---
title: "Semipartial Correlation Derivation"
output: html_document
date: "2025-02-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We generated data with a semipartial correlation of: 

```{r}
# generate data
library(ppcor)
set.seed(711)
n = 10000000
y = rnorm(n, mean=50, sd=10)   # Outcome variable
x1 = rnorm(n, mean=10, sd=5)    # Predictor 1
x2 = rnorm(n, mean=20, sd=5)    # Predictor 2
aux = .05*y + 0.5*x1 + 0.3*x2 + rnorm(n, mean=0, sd=2)  # Auxiliary variable

data = data.frame(y, x1, x2, aux)

# Step 1: Regress the auxiliary variable on other predictors and obtain residuals
aux_resid = resid(lm(aux ~ x1 + x2, data=data))

# Step 2: Compute the correlation between residuals and the outcome
spr_lm = cor(aux_resid, data$y)
print(round(spr_lm,3))
```


To compute the **true semipartial correlation** between the auxiliary variable (AUX) and $Y$, controlling for $X_1$ and $X_2$, we use the following steps.


---

### **Step 1: Define the Auxiliary Variable Equation**
We start with the given equation:

$$
\text{AUX} = 0.05 Y + 0.5 X_1 + 0.3 X_2 + \varepsilon_{\text{aux}}
$$

where $\varepsilon_{\text{aux}} \sim N(0, \sigma_{\text{aux}}^2)$ and $\sigma_{\text{aux}} = 2$.

Also, the standard deviations of the predictors are:

- $\sigma_Y = 10$ (since $Y \sim N(50, 10^2)$)
- $\sigma_{X_1} = 5$ (since $X_1 \sim N(10, 5^2)$)
- $\sigma_{X_2} = 5$ (since $X_2 \sim N(20, 5^2)$)

We assume $X_1, X_2,$ and $Y$ are independent.

---

### **Step 2: Compute the Total Correlation Between AUX and Y**
The total variance of AUX is:

$$
\text{Var}(\text{AUX}) = (0.05)^2 \sigma_Y^2 + (0.5)^2 \sigma_{X_1}^2 + (0.3)^2 \sigma_{X_2}^2 + \sigma_{\text{aux}}^2
$$

$$
= (0.05)^2 (100) + (0.5)^2 (25) + (0.3)^2 (25) + 2^2
$$

$$
= 0.25 + 6.25 + 2.25 + 4 = 12.75
$$

The covariance between AUX and $Y$ is:

$$
\text{Cov}(\text{AUX}, Y) = 0.05 \cdot \sigma_Y^2 = 0.05 \times 100 = 5
$$

Thus, the total Pearson correlation between AUX and $Y$ is:

$$
r_{\text{AUX}, Y} = \frac{\text{Cov}(\text{AUX}, Y)}{\sigma_{\text{AUX}} \sigma_Y} = \frac{5}{\sqrt{12.75} \times 10}
$$

$$
= \frac{5}{3.57 \times 10} = \frac{5}{35.7} \approx 0.14
$$

---

### **Step 3: Compute the Semipartial Correlation**
To obtain the semipartial correlation, we remove the part of AUX that is linearly predicted by $X_1 $ and $X_2 $.

1. **Regress AUX on $X_1$ and $X_2$:**
   $$
   \hat{\text{AUX}} = 0.5 X_1 + 0.3 X_2
   $$

   The variance of this predicted part is:

   $$
   \text{Var}(\hat{\text{AUX}}) = (0.5)^2 \sigma_{X_1}^2 + (0.3)^2 \sigma_{X_2}^2
   $$

   $$
   = (0.5)^2 (25) + (0.3)^2 (25) = 6.25 + 2.25 = 8.5
   $$

2. **Residual Variance of AUX (after regressing on $X_1$ and $X_2$):**
   $$
   \text{Var}(\text{AUX}_{\perp X_1, X_2}) = \text{Var}(\text{AUX}) - \text{Var}(\hat{\text{AUX}})
   $$

   $$
   = 12.75 - 8.5 = 4.25
   $$

3. **Compute the Semipartial Correlation:**
   $$
   r_{\text{AUX}_{\perp X_1, X_2}, Y} = \frac{\text{Cov}(\text{AUX}, Y)}{\sigma_Y \sqrt{\text{Var}(\text{AUX}_{\perp X_1, X_2})}}
   $$

   $$
   = \frac{5}{10 \times \sqrt{4.25}}
   $$

   $$
   = \frac{5}{10 \times 2.06} = \frac{5}{20.6} \approx 0.243
   $$

---

### **Final Answer:**
$$
r_{\text{AUX}_{\perp X_1, X_2}, Y} \approx 0.243
$$


### **Answer from simulation:**

```{r, echo=TRUE}
print(round(spr_lm,3))
```