---
title: "Bayesian Analysis with Missing Data"
output: html_document
---

This document demonstrates Bayesian analysis techniques for handling missing data, comparing different model specifications using both `lavaan` (for structural equation modeling) and `R2jags` (for Bayesian modeling with JAGS).

## Setup

First, we load the necessary R packages.

```{r, echo=TRUE}
if (!require(mvtnorm)){
  install.packages("mvtnorm")
}
library(mvtnorm)

haspackage = require("lavaan")
if (haspackage==FALSE){
  install.packages("lavaan")
}
library(lavaan)

haspackage = require("R2jags")
if (haspackage==FALSE){
  install.packages("R2jags")
}
library(R2jags)
```

#### Data Simulation

We simulate a dataset with two groups (men and women) and several variables. Missingness is introduced in cc, perf, and use based on logistic models.

```{r, echo=TRUE}
set.seed(20250223)

Nmen = 121
Nwomen = 229

MeanMEN = c(5.081333347, 9.723030258, 51.94404048, 52.66452548, 34.04606078, 77.06181845, 14.75389904)
MeanWOMEN = c(4.80418631, 10.60486174, 50.34834542, 48.13359134, 30.61321679, 71.77082955, 13.75449003)

Corr = matrix(c(
  1.0,  .15, .12, .48, .44, .47, .44,
  .15, 1.0, .06, .25, .20, .23, .23,
  .12, .06, 1.0, .40, .32, .19, .14,
  .48, .25, .40, 1.0, .87, .61, .54,
  .44, .20, .32, .87, 1.0, .56, .51,
  .47, .23, .19, .61, .56, 1.0, .70,
  .44, .23, .14, .54, .51, .70, 1.0
), nrow = 7, byrow = TRUE)

SD = c(1.2, 6.0, 15.2, 16.6, 10.9, 10.5, 2.8)
SDdiag = diag(SD)

Cov = SDdiag %*% Corr %*% SDdiag

xmen = rmvnorm(Nmen, MeanMEN, Cov)
xwomen = rmvnorm(Nwomen, MeanWOMEN, Cov)

xmen = cbind(0, xmen)
xwomen = cbind(1, xwomen)

allx = as.data.frame(rbind(xmen, xwomen))
names(allx) = c("female", "hsl", "cc", "use", "msc", "mas", "mse", "perf")

# Missingness in CC
M_CC_intercept = -3
M_CC_female = 1
M_CC_msc = .01
M_CC_mas = .01
M_CC_mse = .01

missingCCmodelLogit = 
  M_CC_intercept +
  M_CC_female*allx[,"female"] +
  M_CC_msc*allx[,"msc"] +
  M_CC_mas*allx[,"mas"] +
  M_CC_mse*allx[,"mse"]

missingCCmodelProb = exp(missingCCmodelLogit)/(1+exp(missingCCmodelLogit))
makeMissingCC = which(runif(Nmen+Nwomen) < missingCCmodelProb)
allx$cc[makeMissingCC] = NA

# Missingness in PERF
M_perf_intercept = -3
M_perf_female = .5
M_perf_msc = .001
M_perf_mas = .02
M_perf_mse = .01

missingPERFmodelLogit = 
  M_perf_intercept +
  M_perf_female*allx[,"female"] +
  M_perf_msc*allx[,"msc"] +
  M_perf_mas*allx[,"mas"] +
  M_perf_mse*allx[,"mse"]

missingPERFmodelProb = exp(missingPERFmodelLogit)/(1+exp(missingPERFmodelLogit))
makeMissingPerf = which(runif(Nmen+Nwomen) < missingPERFmodelProb)
allx$perf[makeMissingPerf] = NA

# Missingness in USE
M_use_intercept = -3
M_use_female = .5
M_use_msc = .001
M_use_mas = .02
M_use_mse = .01

missingUSEmodelLogit = 
  M_use_intercept +
  M_use_female*allx[,"female"] +
  M_use_msc*allx[,"msc"] +
  M_use_mas*allx[,"mas"] +
  M_use_mse*allx[,"mse"]

missingUSEmodelProb = exp(missingUSEmodelLogit)/(1+exp(missingUSEmodelLogit))
makeMissingUse = which(runif(Nmen+Nwomen) < missingUSEmodelProb)
allx$use[makeMissingUse] = NA

data02 = as.data.frame(allx)

# Center CC
par(mfrow = c(1,2))
boxplot(data02$cc, main="Boxplot of College Experience (CC)")
hist(data02$cc, main = "Histogram of College Experience (CC)", xlab = "CC")
par(mfrow = c(1,1))

data02$cc10 = data02$cc - 10

# Interaction term
data02$femXcc10 = data02$female*data02$cc10

newData = data02[c("perf", "use", "female", "cc10")]

# Check for observations missing on all variables
allNA = apply(newData, 1, function(x) all(is.na(x)))
which(allNA)
```


### Model 1: Empty Model

**Model Description** 
This model examines the basic relationships between perf and use without any predictors. In the lavaan specification, perf and use are predicted by their intercepts, and their variances and covariance are estimated. The JAGS model model01a.jags.syntax mirrors this, estimating the means and the covariance matrix of perf and use.

#### lavaan Specification

```{r, echo=TRUE}
model01.lavaan.syntax = "

#Means:
perf ~ 1
use  ~ 1

#Variances:
perf ~~ perf
use  ~~ use

#Covariance:
perf ~~ use

"
model01.lavaan.fit = lavaan(model01.lavaan.syntax, data=data02, estimator = "MLR", mimic = "MPLUS")
summary(model01.lavaan.fit, standardized=TRUE, fit.measures=TRUE)
```



#### R2jags Specification (model01a)
```{r, echo=TRUE}
model01a.jags.syntax <- function(){

    # Priors for means
    mu.perf ~ dnorm(0, 0.001)
    mu.use ~ dnorm(0, 0.001)

    # Prior for inverse covariance matrix (precision matrix)
    Omega[1:2, 1:2] ~ dwish(R, df)
    Sigma <- inverse(Omega) # Covariance matrix

    # Likelihood
    for (i in 1:n) {
      Y[i, 1:2] ~ dmnorm(mu, Omega[,])
    }

    # Means vector
    mu[1] <- mu.perf
    mu[2] <- mu.use

    # Derived parameters (variances and covariance)
    variance.perf <- Sigma[1, 1]
    variance.use <- Sigma[2, 2]
    covariance <- Sigma[1, 2]
    correlation <- covariance / (sqrt(variance.perf) * sqrt(variance.use))
}

model01a.jags.data = list(
  Y = data02[c("perf","use")],
  n = nrow(data02),
  df = 2,
  R = diag(2)
) 
  
model01a.jags.params = c("mu.perf", "mu.use", "variance.perf", "variance.use", "covariance", "correlation")
# 
# model01.jags.samples = jags(
#   data = model01a.jags.data,
#   parameters.to.save = model01a.jags.params,
#   model.file = model01a.jags.syntax,
#   n.chains = 5,
#   n.iter = 6000,
#   n.burnin = 1000,
#   n.thin = 1,
# )
```

#### R2jags Specification (model01b)

model01a.jags.syntax fails because multivariate distributions cannot contain missing data. model01b.jags.syntax uses complete cases only.

```{r, echo=TRUE}
data03 = data02[c("use", "perf")]
data03 = data03[complete.cases(data03),]

model01b.jags.data = list(
  Y = data03,
  n = nrow(data03),
  df = 2,
  R = diag(2)
) 

model01b.jags.samples = jags(
  data = model01b.jags.data,
  parameters.to.save = model01a.jags.params,
  model.file = model01a.jags.syntax,
  n.chains = 4,
  n.iter = 6000,
  n.burnin = 1000,
  n.thin = 1,
)

model01b.jags.samples
plot(as.mcmc(model01b.jags.samples))
```

#### lavaan Specification (model01c)

model01c.lavaan.syntax uses a factored regression approach to handle missing data, perf ~ use and use ~ 1.

```{r, echo=TRUE}
model01c.lavaan.syntax = "

# exogenous mean:
use ~ 1

# exogenous variance:
use ~~ use

# regression:
perf ~ 1 + use

# residual variance:
perf ~~ perf
"

model01c.lavaan.fit = lavaan(model01c.lavaan.syntax, data=data02, estimator = "MLR", mimic = "MPLUS")
summary(model01c.lavaan.fit, standardized=TRUE, fit.measures=TRUE)
```

#### R2jags Specification (model01c)

model01c.jags.syntax mirrors the factored regression approach of model01c.lavaan.syntax in JAGS.

```{r, echo=TRUE}
model01c.jags.syntax <- function(){
  
  # model for use: -----------------
  
  for (obs in 1:N){
    # use mean model: 
    muUse[obs] <- use.intercept
    
    # use model data distribution
    use[obs] ~ dnorm(muUse[obs], tauUse)  
  }
  
  # use prior distributions
  use.intercept ~ dnorm(use.intercept.mean0, use.intercept.precision0)
  tauUse ~ dgamma(use.tau.shape0, use.tau.rate0)
  
  # use variance
  sigma2Use <- 1/tauUse
  
  # model for perf: -----------------
  
  # perf model data distribution
  for (obs in 1:N){
    # perf mean model
    
    muPerf[obs] <- perf.intercept + perf.use*use[obs]
    perf[obs] ~ dnorm(muPerf[obs], tauPerf)    
  }
  

  # perf prior distributions
  perf.intercept ~ dnorm(perf.intercept.mean0, perf.intercept.precision0)
  perf.use ~ dnorm(perf.use.mean0, perf.use.precision0)
  tauPerf ~ dgamma(perf.tau.shape0, perf.tau.rate0)
  
  # perf residual variance
  sigma2Perf <- 1/tauPerf
  
  
}

model01c.jags.data = list(
  use = data02$use,
  perf = data02$perf,
  N = nrow(data02),
  use.intercept.mean0 = 0,
  use.intercept.precision0 = 1/1000,
  use.tau.shape0 = 1,
  use.tau.rate0 = 1,
  perf.intercept.mean0 = 0,
  perf.intercept.precision0 = 1/1000,
  perf.use.mean0 = 0,
  perf.use.precision0 = 1/1000,
  perf.tau.shape0 = 1,
  perf.tau.rate0 = 1
) 

model01c.jags.params = c(
  "use.intercept", 
  "sigma2Use", 
  "perf.intercept", 
  "perf.use", 
  "sigma2Perf",
  "perf[3]",
  "use[8]"
)

model01c.jags.samples = jags(
  data = model01c.jags.data,
  parameters.to.save = model01c.jags.params,
  model.file = model01c.jags.syntax,
  n.chains = 4,
  n.iter = 6000,
  n.burnin = 1000,
  n.thin = 1
)

model01c.jags.samples
plot(as.mcmc(model01c.jags.samples))
```

###  Model 2: Adding Predictors

**Model Description**

Model 2 expands on Model 1 by adding female and cc10 as predictors for both perf and use. The lavaan model estimates the coefficients for these predictors. The JAGS model mirrors this, including priors for the coefficients of female and cc10.

#### lavaan Specification (model02a)

```{r, echo=TRUE}
data02 = data02
data02$female[sample(x = 1:nrow(data02), replace = TRUE, size = .1*nrow(data02) )] = NA

model02a.lavaan.syntax = "

#Means:
perf ~ 1 + female + cc10
use  ~ 1 + female + cc10

#Variances:
perf ~~ perf
use  ~~ use

#Covariance:
perf ~~ use

"

model02.lavaan.fit = lavaan(model02a.lavaan.syntax, data=data02, estimator = "MLR", mimic = "MPLUS")
summary(model02.lavaan.fit, standardized=TRUE, fit.measures=TRUE)
```

#### R2jags Specification (model02a)

```{r, echo=TRUE}
model02a.jags.syntax = function(){
  
  # model for use: -----------------
  
  for (obs in 1:N){
    # use mean model: 
    muUse[obs] <- use.intercept + use.female*female[obs] + use.cc10*cc10[obs]
    
    # use model data distribution
    use[obs] ~ dnorm(muUse[obs], tauUse)  
  }
  
  # use prior distributions -- empirical prior for regression parameters
  use.intercept ~ dnorm(useMean0, usePrecision0)
  use.female ~ dnorm(useMean0, usePrecision0)
  use.cc10 ~ dnorm(useMean0, usePrecision0)
  
  useMean0 ~ dnorm(0, 1/1000)
  usePrecision0 ~ dgamma(1, 1)
  
  tauUse ~ dgamma(use.tau.shape0, use.tau.rate0)
  
  # use variance
  sigma2Use <- 1/tauUse
  
  # model for perf: -----------------
  
  # perf model data distribution
  for (obs in 1:N){
    # perf mean model
    
    muPerf[obs] <- perf.intercept + perf.female*female[obs] + perf.cc10*cc10[obs]
    perf[obs] ~ dnorm(muPerf[obs], tauPerf)    
  }
  

  # perf prior distributions
  perf.intercept ~ dnorm(perfMean0, perfPrecision0)
  perf.female ~ dnorm(perfMean0, perfPrecision0)
  perf.cc10 ~ dnorm(perfMean0, perfPrecision0)
  
  perfMean0 ~ dnorm(0, 1/1000)
  perfPrecision0 ~ dgamma(1, 1)
  
  tauPerf ~ dgamma(perf.tau.shape0, perf.tau.rate0)
  
  # perf residual variance
  sigma2Perf <- 1/tauPerf
  

}

# Prepare data for JAGS -- here perf and use are separate variables
model02a.jags.data = list(
  use = data02$use,
  perf = data02$perf,
  female = data02$female,
  cc10 = data02$cc10,
  N = nrow(data02),
  use.tau.shape0 = 1,
  use.tau.rate0 = 1,
  perf.tau.shape0 = 1,
  perf.tau.rate0 = 1
) 

model02a.jags.params = c(
  "use.intercept", 
  "use.female",
  "use.cc10",
  "sigma2Use", 
  "perf.intercept", 
  "perf.female",
  "perf.cc10",
  "sigma2Perf",
  "perf[3]",
  "use[8]"
)

# Run the model using the R2jags package
# model02a.jags.samples = jags(
#   data = model02a.jags.data,
#   parameters.to.save = model02a.jags.params,
#   model.file = model02a.jags.syntax,
#   n.chains = 4,
#   n.iter = 6000,
#   n.burnin = 1000,
#   n.thin = 1
# )
```

#### R2jags Specification (model02b)

model02a.jags.syntax has problems with missing predictor data. model02b.jags.syntax addresses this by including the predictors female and cc10 in the likelihood. Priors are specified for the parameters of the distributions of female and cc10.

```{r, echo=TRUE}
model02b.jags.syntax = function(){
  
  # model for female -------------
  for (obs in 1:N){
    female[obs] ~ dbern(pFemale)  
  }
  
  # priors for female
  pFemale ~ dbeta(1, 1)
  
  # model for cc10
  for (obs in 1:N){
    cc10[obs] ~ dnorm(muCC10, tauCC10)  
  }
  
  
  # priors for cc10
  muCC10 ~ dnorm(0, 1/1000)
  tauCC10 ~ dgamma(1, 1)
  
  # model for use: -----------------
  
  for (obs in 1:N){
    # use mean model: 
    muUse[obs] <- use.intercept + use.female*female[obs] + use.cc10*cc10[obs]
    
    # use model data distribution
    use[obs] ~ dnorm(muUse[obs], tauUse)  
  }
  
  # use prior distributions -- empirical prior for regression parameters
  use.intercept ~ dnorm(useMean0, usePrecision0)
  use.female ~ dnorm(useMean0, usePrecision0)
  use.cc10 ~ dnorm(useMean0, usePrecision0)
  
  useMean0 ~ dnorm(0, 1/1000)
  usePrecision0 ~ dgamma(1, 1)
  useSigma2_0 = 1/usePrecision0
  
  tauUse ~ dgamma(use.tau.shape0, use.tau.rate0)
  
  # use variance
  sigma2Use <- 1/tauUse
  
  # model for perf: -----------------
  
  # perf model data distribution
  for (obs in 1:N){
    # perf mean model
    muPerf[obs] <- perf.intercept + perf.female*female[obs] + perf.cc10*cc10[obs]
    perf[obs] ~ dnorm(muPerf[obs], tauPerf)    
  }
  
  
  # perf prior distributions
  perf.intercept ~ dnorm(perfMean0, perfPrecision0)
  perf.female ~ dnorm(perfMean0, perfPrecision0)
  perf.cc10 ~ dnorm(perfMean0, perfPrecision0)
  
  perfMean0 ~ dnorm(0, 1/1000)
  perfPrecision0 ~ dgamma(1, 1)
  perfSigma2_0 = 1/perfPrecision0
  
  tauPerf ~ dgamma(perf.tau.shape0, perf.tau.rate0)
  
  # perf residual variance
  sigma2Perf <- 1/tauPerf
  
  
}

# Prepare data for JAGS -- here perf and use are separate variables
model02b.jags.data = list(
  use = data02$use,
  perf = data02$perf,
  female = data02$female,
  cc10 = data02$cc10,
  N = nrow(data02),
  use.tau.shape0 = 1,
  use.tau.rate0 = 1,
  perf.tau.shape0 = 1,
  perf.tau.rate0 = 1
) 

model02b.jags.params = c(
  "pFemale",
  "muCC10",
  "tauCC10",
  "use.intercept", 
  "use.female",
  "use.cc10",
  "useMean0",
  "useSigma2_0",
  "sigma2Use", 
  "perf.intercept", 
  "perf.female",
  "perf.cc10",
  "sigma2Perf",
  "perfMean0",
  "perfSigma2_0",
  "perf[3]",
  "use[8]"
)

# Run the model using the R2jags package
model02b.jags.samples = jags(
  data = model02b.jags.data,
  parameters.to.save = model02b.jags.params,
  model.file = model02b.jags.syntax,
  n.chains = 4,
  n.iter = 6000,
  n.burnin = 1000,
  n.thin = 1
)

model02b.jags.samples 
```


### Model 3: Adding Interactions

**Model Description** 

Model 3 adds interaction terms (cc10 x female) to both the perf and use equations in the JAGS model. This allows the effect of cc10 to vary depending on female, and vice-versa. There is no equivalent lavaan model provided in the code.

#### R2jags Specification (model03a)

```{r, echo=TRUE}
model03a.jags.syntax = function(){
  
  # model for female -------------
  for (obs in 1:N){
    female[obs] ~ dbern(pFemale)  
  }
  
  # priors for female
  pFemale ~ dbeta(1, 1)
  
  # model for cc10
  for (obs in 1:N){
    cc10[obs] ~ dnorm(muCC10, tauCC10)  
  }
  
  
  # priors for cc10
  muCC10 ~ dnorm(0, 1/1000)
  tauCC10 ~ dgamma(1, 1)
  
  # model for use: -----------------
  
  for (obs in 1:N){
    # use mean model: 
    muUse[obs] <- use.intercept + use.female*female[obs] + use.cc10*cc10[obs] + 
      use.cc10xfemale*cc10[obs]*female[obs]
    
    # use model data distribution
    use[obs] ~ dnorm(muUse[obs], tauUse)  
  }
  
  # use prior distributions -- empirical prior for regression parameters
  use.intercept ~ dnorm(useMean0, usePrecision0)
  use.female ~ dnorm(useMean0, usePrecision0)
  use.cc10 ~ dnorm(useMean0, usePrecision0)
  use.cc10xfemale ~ dnorm(useMean0, usePrecision0)
  
  useMean0 ~ dnorm(0, 1/1000)
  usePrecision0 ~ dgamma(1, 1)
  useSigma2_0 = 1/usePrecision0
  
  tauUse ~ dgamma(use.tau.shape0, use.tau.rate0)
  
  # use variance
  sigma2Use <- 1/tauUse
  
  # model for perf: -----------------
  
  # perf model data distribution
  for (obs in 1:N){
    # perf mean model
    muPerf[obs] <- perf.intercept + perf.female*female[obs] + perf.cc10*cc10[obs] + 
      perf.cc10xfemale*cc10[obs]*female[obs]
    perf[obs] ~ dnorm(muPerf[obs], tauPerf)    
  }
  
  
  # perf prior distributions
  perf.intercept ~ dnorm(perfMean0, perfPrecision0)
  perf.female ~ dnorm(perfMean0, perfPrecision0)
  perf.cc10 ~ dnorm(perfMean0, perfPrecision0)
  perf.cc10xfemale ~ dnorm(perfMean0, perfPrecision0)
  
  perfMean0 ~ dnorm(0, 1/1000)
  perfPrecision0 ~ dgamma(1, 1)
  perfSigma2_0 = 1/perfPrecision0
  
  tauPerf ~ dgamma(perf.tau.shape0, perf.tau.rate0)
  
  # perf residual variance
  sigma2Perf <- 1/tauPerf
  
  
}

# Prepare data for JAGS -- here perf and use are separate variables
model03a.jags.data = list(
  use = data02$use,
  perf = data02$perf,
  female = data02$female,
  cc10 = data02$cc10,
  N = nrow(data02),
  use.tau.shape0 = 1,
  use.tau.rate0 = 1,
  perf.tau.shape0 = 1,
  perf.tau.rate0 = 1
) 

model03a.jags.params = c(
  "pFemale",
  "muCC10",
  "tauCC10",
  "use.intercept", 
  "use.female",
  "use.cc10",
  "use.cc10xfemale",
  "useMean0",
  "useSigma2_0",
  "sigma2Use", 
  "perf.intercept", 
  "perf.female",
  "perf.cc10",
  "perf.cc10xfemale",
  "sigma2Perf",
  "perfMean0",
  "perfSigma2_0",
  "perf[3]",
  "use[8]"
)

# Run the model using the R2jags package
model03a.jags.samples = jags(
  data = model03a.jags.data,
  parameters.to.save = model03a.jags.params,
  model.file = model03a.jags.syntax,
  n.chains = 4,
  n.iter = 6000,
  n.burnin = 1000,
  n.thin = 1
)

model03a.jags.samples 
```

### Model 4: Adding Auxiliary Variables

**Model Description**

Model 4 expands the model to include additional auxiliary variables (hsl, msc, mas, and mse). Critically, because these auxiliary variables also have missing data, the JAGS model now includes a joint model for all the variables with missingness (female, cc10, hsl, msc, mas, and mse).This joint model is factored into a series of regressions.


#### R2jags Specification (model04a)
```{r, echo=TRUE}
model04a.jags.syntax = function(){
  
  # model for hsl ---------------

  for (obs in 1:N){
    hsl[obs] ~ dnorm(muHSL, tauHSL)  
  }
  
  # priors for hsl
  muHSL ~ dnorm(0, 1/1000)
  tauHSL ~ dgamma(1, 1)
  
  # model for msc ---------------
  
  for (obs in 1:N){
    mscMean[obs] <- msc.intercept + msc.hsl*hsl[obs]
    msc[obs] ~ dnorm(mscMean[obs], tauMSC)
  }
  
  # priors for msc
  msc.intercept ~ dnorm(muMSC0, tauMSC0)
  msc.hsl ~ dnorm(muMSC0, tauMSC0)
  muMSC0 ~ dnorm(0, 1/1000)
  tauMSC0 ~ dgamma(1, 1)
  tauMSC ~ dgamma(1, 1)
  
  # model for mas ---------------
  
  for (obs in 1:N){
    masMean[obs] <- mas.intercept + mas.hsl*hsl[obs] + mas.msc*msc[obs]
    mas[obs] ~ dnorm(masMean[obs], tauMAS)
  }
  
  # priors for mas
  mas.intercept ~ dnorm(muMAS0, taoMAS0)
  mas.hsl ~ dnorm(muMAS0, taoMAS0)
  mas.msc ~ dnorm(muMAS0, taoMAS0)
  muMAS0 ~ dnorm(0, 1/1000)
  taoMAS0 ~ dgamma(1, 1)
  tauMAS ~ dgamma(1, 1)
  
  # model for mse ---------------
  
  for (obs in 1:N){
    mseMean[obs] <- mse.intercept + mse.hsl*hsl[obs] + mse.msc*msc[obs] + mse.mas*mas[obs]
    mse[obs] ~ dnorm(mseMean[obs], tauMSE)
  }
  
  # priors for mse
  mse.intercept ~ dnorm(muMSE0, tauMSE0)
  mse.hsl ~ dnorm(muMSE0, tauMSE0)
  mse.msc ~ dnorm(muMSE0, tauMSE0)
  mse.mas ~ dnorm(muMSE0, tauMSE0)
  muMSE0 ~ dnorm(0, 1/1000)
  tauMSE0 ~ dgamma(1, 1)
  tauMSE ~ dgamma(1, 1)
  
    
  # model for female -------------
  for (obs in 1:N){
    pFemale[obs] = ilogit(female.intercept + female.hsl*hsl[obs] + 
                            female.msc*msc[obs] + female.mas*mas[obs] + female.mse*mse[obs])
    female[obs] ~ dbern(pFemale[obs])
  }
  
  # priors for female
  female.intercept ~ dnorm(muFemale0, tauFemale0)
  female.hsl ~ dnorm(muFemale0, tauFemale0)
  female.msc ~ dnorm(muFemale0, tauFemale0)
  female.mas ~ dnorm(muFemale0, tauFemale0)
  female.mse ~ dnorm(muFemale0, tauFemale0)
  muFemale0 ~ dnorm(0, 1/1000)
  tauFemale0 ~ dgamma(1, 1)
  
  
  # model for cc10
  for (obs in 1:N){
    muCC10[obs] = cc10.intercept + cc10.hsl*hsl[obs] + cc10.msc*msc[obs] + 
      cc10.mas*mas[obs] + cc10.mse*mse[obs]
    cc10[obs] ~ dnorm(muCC10[obs], tauCC10)  
  }
  
  
  # priors for cc10
  cc10.intercept ~ dnorm(muCC10_0, tauCC10_0)
  cc10.hsl ~ dnorm(muCC10_0, tauCC10_0)
  cc10.msc ~ dnorm(muCC10_0, tauCC10_0)
  cc10.mas ~ dnorm(muCC10_0, tauCC10_0)
  cc10.mse ~ dnorm(muCC10_0, tauCC10_0)
  muCC10_0 ~ dnorm(0, 1/1000)
  tauCC10_0 ~ dgamma(1, 1)
  
  tauCC10 ~ dgamma(1, 1)
  
  # model for use: -----------------
  
  for (obs in 1:N){
    # use mean model: 
    muUse[obs] <- use.intercept + use.female*female[obs] + use.cc10*cc10[obs] + 
      use.cc10xfemale*cc10[obs]*female[obs] + use.hsl*hsl[obs] + use.msc*msc[obs] + 
      use.mas*mas[obs] + use.mse*mse[obs]
    
    # use model data distribution
    use[obs] ~ dnorm(muUse[obs], tauUse)  
  }
  
  # use prior distributions -- empirical prior for regression parameters
  use.intercept ~ dnorm(useMean0, usePrecision0)
  use.female ~ dnorm(useMean0, usePrecision0)
  use.cc10 ~ dnorm(useMean0, usePrecision0)
  use.cc10xfemale ~ dnorm(useMean0, usePrecision0)
  use.hsl ~ dnorm(useMean0, usePrecision0)
  use.msc ~ dnorm(useMean0, usePrecision0)
  use.mas ~ dnorm(useMean0, usePrecision0)
  use.mse ~ dnorm(useMean0, usePrecision0)
  
  useMean0 ~ dnorm(0, 1/1000)
  usePrecision0 ~ dgamma(1, 1)
  useSigma2_0 = 1/usePrecision0
  
  tauUse ~ dgamma(use.tau.shape0, use.tau.rate0)
  
  # use variance
  sigma2Use <- 1/tauUse
  
  # model for perf: -----------------
  
  # perf model data distribution
  for (obs in 1:N){
    # perf mean model
    muPerf[obs] <- perf.intercept + perf.female*female[obs] + perf.cc10*cc10[obs] + 
      perf.cc10xfemale*cc10[obs]*female[obs] + perf.hsl*hsl[obs] + perf.msc*msc[obs] +
      perf.mas*mas[obs] + perf.mse*mse[obs]
    perf[obs] ~ dnorm(muPerf[obs], tauPerf)    
  }
  
  
  # perf prior distributions
  perf.intercept ~ dnorm(perfMean0, perfPrecision0)
  perf.female ~ dnorm(perfMean0, perfPrecision0)
  perf.cc10 ~ dnorm(perfMean0, perfPrecision0)
  perf.cc10xfemale ~ dnorm(perfMean0, perfPrecision0)
  perf.hsl ~ dnorm(perfMean0, perfPrecision0)
  perf.msc ~ dnorm(perfMean0, perfPrecision0)
  perf.mas ~ dnorm(perfMean0, perfPrecision0)
  perf.mse ~ dnorm(perfMean0, perfPrecision0)
  
  perfMean0 ~ dnorm(0, 1/1000)
  perfPrecision0 ~ dgamma(1, 1)
  perfSigma2_0 = 1/perfPrecision0
  
  tauPerf ~ dgamma(perf.tau.shape0, perf.tau.rate0)
  
  # perf residual variance
  sigma2Perf <- 1/tauPerf
  
  
}

# Prepare data for JAGS -- here perf and use are separate variables
model04a.jags.data = list(
  use = data02$use,
  perf = data02$perf,
  female = data02$female,
  cc10 = data02$cc10,
  hsl = data02$hsl,
  msc = data02$msc,
  mas = data02$mas,
  mse = data02$mse,
  N = nrow(data02),
  use.tau.shape0 = 1,
  use.tau.rate0 = 1,
  perf.tau.shape0 = 1,
  perf.tau.rate0 = 1
) 

model04a.jags.params = c(
  "muHSL",
  "tauHSL",
  "msc.intercept",
  "msc.hsl",
  "tauMSC",
  "mas.intercept",
  "mas.hsl",
  "mas.msc",
  "tauMAS",
  "mse.intercept",
  "mse.hsl",
  "mse.msc",
  "mse.mas",
  "tauMSE",
  "female.intercept",
  "female.hsl",
  "female.msc",
  "female.mas",
  "female.mse",
  "cc10.intercept",
  "cc10.hsl",
  "cc10.msc",
  "cc10.mas",
  "cc10.mse",
  "muCC10_0",
  "tauCC10_0",
  "use.intercept", 
  "use.female",
  "use.cc10",
  "use.cc10xfemale",
  "use.hsl",
  "use.msc",
  "use.mas",
  "use.mse",
  "useMean0",
  "useSigma2_0",
  "sigma2Use", 
  "perf.intercept", 
  "perf.female",
  "perf.cc10",
  "perf.cc10xfemale",
  "perf.hsl",
  "perf.msc",
  "perf.mas",
  "perf.mse",
  "sigma2Perf",
  "perfMean0",
  "perfSigma2_0",
  "perf[3]",
  "use[8]"
)

# Run the model using the R2jags package
model04a.jags.samples = jags(
  data = model04a.jags.data,
  parameters.to.save = model04a.jags.params,
  model.file = model04a.jags.syntax,
  n.chains = 4,
  n.iter = 6000,
  n.burnin = 1000,
  n.thin = 1
)

model04a.jags.samples 
```
